---
title: "Fish summarizer function"
author: "Anela Akiona, Stuart Sandin, Brian Zgliczynski"
output:
  word_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This Rmd is an R markdown function that calculates density and biomass of fish transect count data, using user-provided length-weight parameters, and produces summary files and plots. 

## The function inputs are:
- fishdata -- the csv file containing the field observation data of fish counts and lengths.
- LWfile -- the csv file containing the fish life history parameters.
  - *NOTE: this function will not work if there are species missing length-weight parameters*
- filename -- unique island/site name to identify output files.
- sp_data -- the column name which contains the species names for the fish data file.
- sp_lw -- the column name which contains the species names for the length-weight parameters. These values *must* match the values in sp_data.
- integrate -- if TRUE, uses integral method to calculate biomass. If FALSE, uses the given length.
- size_correct -- number to be subtracted from the total length of each fish prior to biomass calculation. Cannot be used if integrate = TRUE.
- bin -- bin size used for length measurements (cm). Can either be a single number or a vector of numbers.
- max_size -- maximum size to which biomass should be calculated when bin is a single number (cm).
- size_threshold -- fish length at which belt transect width changes (cm)
- width_below -- width of transect for fish smaller than size_threshold (m).
- width_above -- width of transect for fish larger than size_threshold (m).
- transect_length -- length of transect (m).


## Outputs
Data files:
-	fish_summary_station_'filename'.csv
  -	Biomass and density for each station.
-	fish_summary_group_'filename'.csv
  -	Biomass and density by trophic group at each station.
-	fish_summary_species_'filename'.csv
  -	Biomass and density by species at each station.
Plots:
-	fish_summary_density_'filename'.png
  -	Stacked bar chart of density (#/m2) by trophic group at each station.
-	fish_summary_biomass_'filename'.png
  -	Stacked bar chart of biomass (g/m2) by trophic group at each station.

# USAGE NOTES:
Run the below chunk (lines 49-541) to load the fish summarizer function. Then, in a separate chunk, run fish_summary() with the required arguments (fishdata, LWfile, and filename). Option to alter the other arguments if you wish to change the defaults.

```{r Function - DO NOT ALTER}
fish_summary <- function(fishdata, LWfile, filename, 
                         sp_data = "Species", sp_lw = "NewName",
                         integrate = TRUE, 
                         size_correct = 0,
                         bin = 5, max_size = 300, size_threshold = 20, 
                         width_below = 2, width_above = 4, transect_length = 25) {


# Load packages ####
library(Hmisc)
library(gdata)
library(readxl)
library(ggplot2)
library(tidyr)
library(dplyr)
library(tibble)
library(fishualize)


#### Import data files and QC #####
  
# Read in "fish", the raw survey data
### NOTE: data needs to be in csv format in order for columns to be recognized as factors
fish <- read.csv(fishdata, header = TRUE)

# Remove any NA rows/columns added by R
fish <- fish[colSums(!is.na(fish)) > 0] # cols
fish <- fish[!is.na(fish$Size), ]       # rows


# Read in "LW", the species-specific lookup table, with length-weight data and trophic (or other) classifications
LW <- read.csv(LWfile, header = TRUE)

n <- length(fish[, 1])


# Brief routine to check if all species listed in 'fish' are contained within 'LW'
### ADDED 28 Sept 2015

# Needs to be up here or sometimes an error will pop up if there are spp missing
spp <- levels(drop.levels(as.factor(fish[ , sp_data])))
n_spp <- length(spp)


#################################################
#
# QC routines ####
#
#################################################
### Length-weight parameters ####
ii <- 1

for(i in 1:n) {
  
  temp_spp <- fish[i, sp_data]
  jj <- match(temp_spp, LW[ , sp_lw], nomatch = NA)
  
  if(is.na(jj)==TRUE && ii==1) {
    missing_spp = temp_spp
    ii <- 2
  } 
  
  if(is.na(jj)==TRUE && ii>1) {
    temp_missing <- match(temp_spp,spp[missing_spp], nomatch=NA)
    
    if(is.na(temp_missing)==TRUE) {
      missing_spp <- c(missing_spp,temp_spp)
      ii <- ii+1
    }
  }
  
} #i

#########

# List of the species codes for the missing species, to allow additions to the LW core table
### NOTE -- if response of below is "Error: object 'missing_spp' not found' comes up, then the LW table is complete. Good news!

if(exists("missing_spp") == TRUE){
  warning("Some species are missing length-weight parameters.")
  print(unique(missing_spp))

} else {
  message("All species have length-weight parameters.")

}


### Check fish column names ####
expected_cols_fish <- c("Data_Type", "Date", "Transect", "Diver", "Station", "Number", "Size")

check_fish_cols <- function(df) {
  actual_cols <- colnames(df)
  
  missing_cols <- setdiff(expected_cols_fish, actual_cols)
  
  if (length(missing_cols) == 0) {
    message("All required columns are present in the fish data.")
  } else {
    warning("Some fish data columns are named incorrectly or are missing.")
    cat("Missing columns:\n")
    print(missing_cols)
  }
}

check_fish_cols(fish)

### Check LW column names ####
expected_cols_lw <- c("a_cm", "b_cm", "LTLRat", "Trophic")

# Function to check column names
check_lw_cols <- function(df) {
  actual_cols <- colnames(df)
  
  missing_cols <- setdiff(expected_cols_lw, actual_cols)
  
  if (length(missing_cols) == 0) {
    message("All required columns are present in the length-weight parameter table.")
  } else {
    warning("Some required columns in the length-weight table are missing or are named incorrectly.")
    cat("Missing columns:\n")
    print(missing_cols)
  }
}

check_lw_cols(LW)


#################################################
#
# Beginning of summary routine ####
#
#################################################

# Create parameters from dataset to pass to for loop ####
## Need to use drop.levels to remove blank level that randomly gets added
# sp_data and sp_lw are user-specified column names that identify the species in the respective data files

station <- levels(drop.levels(as.factor(fish$Station)))
n_station <- length(station)

groups <- levels(drop.levels(as.factor(LW$Trophic)))
n_groups <- length(groups)

species <- levels(drop.levels(as.factor(LW[ , sp_lw])))
n_species <- length(species)



# Lookup effort table for the total number of observers x transect x date combinations for each site ####
### This reflects the total effort, in units of observer-specific individual 25m-long transects (two widths, following DeMartini / Friedlander / Sandin)

effort <- array(0, c(n_station), dimnames = list(station))

for (i in 1:n_station) {
  
  effort[i] <- dim(unique(fish[which(fish$Station == station[i]), c("Date", "Diver", "Transect")]))[1]
  
} #i


# Create empty matrices/arrays to store values ####
### Per capita biomass of a fish
pc_biomass <- matrix(nrow = n, ncol = 1) 

# Create empty column in df to populate with biomass of each fish
fish$pcbio <- NA

### Numbers and biomass at each station
numbers <- array(0, c(n_station), dimnames = list(station)) 
biomass <- array(0, c(n_station), dimnames = list(station)) 

### Number and biomass by functional group
numbers_groups <- array(0, c(n_station, n_groups), dimnames = list(station, groups))
biomass_groups <- array(0, c(n_station, n_groups), dimnames = list(station, groups)) 

### Number and biomass by species by station
numbers_species <- array(0, c(n_station, n_species), dimnames = list(station, species)) 
biomass_species <- array(0, c(n_station, n_species), dimnames = list(station, species)) 



#################################################
#
# Create biomass lookup table ####
# By AKA 8/14/24
#
#################################################


# Create new df with desired columns
lookup <- LW %>%
  select(all_of(c("Taxon", sp_lw, "LTLRat", "a_cm", "b_cm", "Family", "Trophic")))

# Define # to start inner loop at (first column that isn't metadata)
start_loop <- ncol(lookup) + 1

### Need to decide if I want to keep this in here
#lookup$Family <- as.factor(lookup$Family) # convert to factor

## Create new columns depending on whether bin is just 1 value or a vector of values
    if (length(bin) == 1){
      new_cols <- as.character(seq(bin, max_size, bin)) # create names for new columns
    }

    if (length(bin) > 1){
      new_cols <- as.character(bin)
      }

# Create and add new columns for each size class
## Starting at bin omits the 0 size class
lookup[new_cols] <- NA

n_lookup <- length(lookup[ , sp_lw]) # define length of outer loop
end_loop <- ncol(lookup) # Define # to end inner loop at (last column of bins)



#################################################
#
# For loop to create lookup table ####
#
#################################################

for (i in 1:n_lookup) {
  
  # Define length-weight relationship W = a(L^b)
  ## L is the total length of the fish (cm)
  ## Gives the mass (g) of the fish at a given L
  
  LWfun <- function(L) {
    
    lookup$a_cm[i] * ((lookup$LTLRat[i] * L)^lookup$b_cm[i]) # have to call the specific a/b/LTLRat for each species
  
  } # LWfun
  
  # Loop through each size bin column
  for (j in start_loop:end_loop) { # excludes metadata columns 
    
    # Use the integral method to calculate mass
    if (integrate == TRUE){
    
      # For first iteration of loop, define lower limit of integration as 0
      if (j == start_loop){
        # Extract column name as numeric to call in integrate function
        size_upper <- as.numeric(colnames(lookup)[j])   # extract column name (size bin)
    
        # Integrate LW function across size class
        integral <- (integrate(LWfun, lower = 0, upper = size_upper))
        
        # Take average of integral value over bin width, assign to df
        lookup[i, j] <- (1/(size_upper)) * integral$value
        
      } # j == start
      
      
      if (j != start_loop){
        # Extract column name as numeric to call in integrate function
        size_upper <- as.numeric(colnames(lookup)[j])   # extract column name (size bin)
        size_lower <- as.numeric(colnames(lookup)[j - 1])   # lower size bin
        
        # Integrate LW function across size class
        integral <- (integrate(LWfun, lower = size_lower, upper = size_upper))
        
        # Take average of integral value over bin width, assign to df
        lookup[i, j] <- (1/(size_upper - size_lower)) * integral$value
     
       } # j != start
      
    } # integrate == TRUE
    
    
    # When not using integral method, use the value of the size bin to calculate mass.
    if (integrate == FALSE){
      
      lookup[i, j] <- LWfun(as.numeric(colnames(lookup)[j]) - size_correct)
    } # integrate == FALSE
    
  } # j
  
} # i     end of lookup table routine




#################################################
#
# Beginning of for loop to calculate density and biomass ####
# Modified by AKA 2/28/20
#
#################################################

# This for loop:
#   (1) Calculates per capita biomass of fish from a row of data
#   (2) Creates matrix of biomass and numbers (density) of fish at each station
#   (3) Creates matrix of biomass and numbers (density) by functional group
#   (4) Creates matrix of biomass and numbers (density) by species by station



for (i in 1:n) { # where n = length(fish[,1])
  if (fish$Data_Type[i] == "QUAN") {
    
    jj <- match(fish$Station[i],station)    
    spp <- match(fish[i, sp_data], LW[ , sp_lw], nomatch = NA)
    temp_trophic <- LW$Trophic[spp]
    temp_spp <- as.character(fish$Species[i])
    
    # Define conversion factor to standardize to units per m^2 
    ## (# of unique person-transects x [4 or 2]m x 25m transect)
    if (fish$Size[i] > size_threshold) {convert = (1 / effort[jj]) * (1 / (width_above * transect_length))
    } else {convert = (1 / effort[jj]) * (1 / (width_below * transect_length))}
    
    ##################
    
    # (1) Calculate the per capita biomass of a fish from a row of data
    
    ## Calculate biomass
    ### W = aL^b
    
    # Match fish size with column of lookup table (fish size) and pull cell value for biomass
    col_size <- match(fish$Size[i], names(lookup))
    pc_biomass[i] <- lookup[spp, col_size]

    ##################
    
    # (2) Create matrix of biomass and numbers (density) of fish at each station
    numbers[jj] <- numbers[jj] + (convert * fish$Number[i])
    biomass[jj] <- biomass[jj] + (convert * (fish$Number[i] * pc_biomass[i]))
    
    ##################
    
    # (3) Create matrix of biomass and numbers (density) by functional group
    numbers_groups[jj, temp_trophic] <- numbers_groups[jj, temp_trophic] + (convert * fish$Number[i])
    biomass_groups[jj, temp_trophic] <- biomass_groups[jj, temp_trophic] + (convert * (fish$Number[i] * pc_biomass[i]))
    
    ##################
    
    # (4) Create matrix of biomass and numbers (density) by species by station
    numbers_species[jj, temp_spp] <- numbers_species[jj, temp_spp] + (convert * fish$Number[i])
    biomass_species[jj, temp_spp] <- biomass_species[jj, temp_spp] + (convert * (fish$Number[i] * pc_biomass[i]))
    
    
  } # if
  
} # for i



#################################
#
# CREATE SUMMARY DATA FRAMES ####
# Added by AKA 3/3/20
#
#################################

# Reformat summary matrices into data frames with columns:
#   Station
#   Trophic_group or Species
#   Density (#/m2)
#   Biomass (g/m2)

#################################

### For biomass and density by STATION

# Convert matrices to tibbles, add column of station names, rename column
station_numb <- as_tibble(numbers) %>% 
  add_column(station, .before = 1)
colnames(station_numb)[2] <- "Density"
  
station_bio <- as_tibble(biomass) %>% 
  add_column(station, .before = 1)
colnames(station_bio)[2] <- "Biomass"

# Combine into one data frame by cols "Trophic_Group", "station"
summary_station <- as.data.frame(inner_join(station_numb, station_bio))


#################################

### For biomass and density by TROPHIC GROUP

# Convert matrices to tibbles, add column of station names, reformat
group_numb <- as_tibble(numbers_groups) %>% 
  add_column(station, .before = 1) %>% 
  gather(key = "Trophic_group", value = "Density", -1)
  
group_bio <- as_tibble(biomass_groups) %>% 
  add_column(station, .before = 1) %>% 
  gather(key = "Trophic_group", value = "Biomass", -1)

# Combine into one data frame by cols "Trophic_Group", "station"
summary_group <- as.data.frame(inner_join(group_numb, group_bio))


#################################

### For biomass and density by SPECIES

# Convert matrices to tibbles, add column of station names, reformat
species_numb <- as_tibble(numbers_species) %>% 
  add_column(station, .before = 1) %>% 
  gather(key = "Species", value = "Density", -1)

species_bio <- as_tibble(biomass_species) %>% 
  add_column(station, .before = 1) %>% 
  gather(key = "Species", value = "Biomass", -1)

# Combine into one data frame by cols "Trophic_Group", "station"
summary_species <- as.data.frame(inner_join(species_numb, species_bio))

# Routine to add Family to df
matches <- match(summary_species$Species, LW$NewName)
summary_species$Family <- LW$Family[matches]

# k <- length(summary_species$Species)
# summary_species$Family <- NA
# 
# for (i in 1:k) {
#   jj <- match(summary_species$Species[i], LW$NewName)
#   summary_species$Family[i] <- LW$Family[jj]
# }

summary_species <- summary_species %>%
  filter(Biomass != 0 & Density != 0) # remove species that weren't observed
summary_species <- summary_species[ , c(1, 2, 5, 3, 4)] # reorder columns



#################################
#
# CREATE OUTPUT FILES ####
# Modified by AKA 3/3/20
#
#################################

# Write csv files
#write.csv(fish, file = paste0("fish_with_biomass_", filename, ".csv"), row.names = FALSE)
write.csv(summary_station, file = paste0("fish_summary_station_", filename, ".csv"), row.names = FALSE)
write.csv(summary_group, file = paste0("fish_summary_group_", filename, ".csv"), row.names = FALSE)
write.csv(summary_species, file = paste0("fish_summary_species_", filename, ".csv"), row.names = FALSE)



#################################
#
# CREATE SUMMARY PLOTS = ####
# Added by AKA 8/14/24
#
#################################


summary_group$Trophic_group <- factor(summary_group$Trophic_group, ordered = TRUE, levels = c("Shark", "Top-predator", "Lower-carnivore", "Planktivore", "Herbivore/Detritivore"))

plotcol <- c("red", "black",  "gray48", "gainsboro", "chartreuse3")

# Abundance
density_plot <- ggplot(summary_group, aes(fill = Trophic_group, y = Density, x = station)) +
  geom_bar(position = "stack", stat = "identity", colour = "black") + # colour add black outline to bar sections
  theme_classic() +
  scale_fill_manual(values = plotcol) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  labs(x = "Station",
    y = bquote('Abundance (#' ~m^-2~')'),
    fill = "Trophic group") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), # rotate x axis labels and shift down
        text = element_text(family = "sans")) +
  ggtitle(filename)

# Biomass
biomass_plot <- ggplot(summary_group, aes(fill = Trophic_group, y = Biomass, x = station)) +
  geom_bar(position = "stack", stat = "identity", colour = "black") +
  theme_classic() +
  scale_fill_manual(values = plotcol) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  labs(x = "Station",
    y = bquote('Biomass (g' ~m^-2~')'),
    fill = "Trophic group") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), # rotate x axis labels and shift down
        text = element_text(family = "sans")) +
  ggtitle(filename)

print(density_plot)
print(biomass_plot)

ggsave(density_plot, filename = paste0("fish_summary_density_", filename, ".png"))
ggsave(biomass_plot, filename = paste0("fish_summary_biomass_", filename, ".png"))

} # fish_summary

```


Example function usage:

```{r}
fish_summary(fishdata = "Fiji_2023-05_QC.csv", 
            LWfile = "Pacific_LW_params.csv",
            filename = "Fiji_2023-05",
            sp_data = "Species", 
            sp_lw = "NewName",
            integrate = FALSE,
            bin = 5,
            max_size = 300,
            size_threshold = 20,
            width_below = 2,
            width_above = 4,
            transect_length = 25)

```
